---
title: "Survival HW3"
author: "David"
date: "Due February 26, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r load_packages, include=FALSE}

library(survival) 
library(flexsurv)
library(survMisc) # for Wilcoxon-Gehan-Breslow test 
source("/Users/david/Documents/GitHub/Survival/Homework 3/getmedianres.R")


```


```{r load_data, include=FALSE}

luke <- read.csv("/Users/david/Documents/GitHub/Survival/Homework 3/ccg803.csv", header=TRUE)
surv.luke <- Surv(time=luke$duration, event=luke$relapse, type="right")

```


<br>
<br>

###Question 1

(a) There is a significant difference in relapse associated with dactinomycin use (p=0.0439). The hazard ratio for relapse comparing those who receive dactinomycin to those who do not is 0.738 (95% CI: 0.5550-0.992). 

```{r fit_coxph1, include=FALSE}

model.1 <- coxph(surv.luke ~ rx, data=luke)
summary(model.1)

```


<br>

(b) There is a significant association between treatment and relapse after controlling for white blood cell count and age using a Cox PH model (p=0.0492). The hazard ratio for relapse comparing those that received dactinomycin to those that did not is 0.744 (95% CI: 0.553-1.000). 

```{r fit_coxph2, include=FALSE}

model.2 <- coxph(surv.luke ~ rx + wbc + age, data=luke)
summary(model.2)

```

<br>

(c) For the sub-population in which the white blood cell (wbc) count is below 10,000, the HR comparing individuals who receive treatment as compared to those who did not is 0.553 (95% CI: 0.368-0.831). For the sub-population in which the wbc count is greater than or equal to 10,000, the HR comparing those who received treatment to those who did not is 1.016 (95% CI: 0.699-1.477). 

```{r fit_cox3, include=FALSE}

#creating interaction variables
luke$wbc_high <- (luke$wbc >= 100)*luke$wbc
luke$wbc_low <- (luke$wbc < 100)*luke$wbc
luke$rx_high <- (luke$wbc >= 100)*luke$rx
luke$rx_low <- (luke$wbc < 100)*luke$rx

#model with interaction
model.3_large <- coxph(surv.luke ~ wbc_high + wbc_low + rx_high + rx_low + age, data = luke)
summary(model.3_large)

model.3_large_est <- exp(coef(model.3_large)["rx_high"])
model.3_large_est

#model without interaction
model.3_small <- coxph(surv.luke ~ rx + wbc_high + wbc_low + age, data=luke)
summary(model.3_small)

#performing test to determine interaction
as.numeric(2 * (logLik(model.3_large) - logLik(model.3_small)))
#test statistic is 5.526



```

<br>

(d) The HR comparing those who received treatment to those who did not, adjusting for wbc, age, and treatment site is 0.710 (95% CI: 0.520-0.969).

```{r fit_cox4, include=FALSE}

model.4 <- coxph(surv.luke ~ rx + wbc + age + strata(institution), data=luke)
summary(model.4)

```


(e) 


```{r plot_curves}

plot(survfit(model.2,
     newdata=data.frame(rx=1, age=5, wbc=45)),
     conf.int=FALSE, lwd=2, col="blue", lty="dashed", 
     xlab="Days", ylab="Survival Probability")
lines(survfit(model.2,
     newdata=data.frame(rx=0, age=5, wbc=45)),
     conf.int=FALSE, lwd=2, col="blue", lty="solid")
lines(survfit(model.2,
     newdata=data.frame(rx=1, age=5, wbc=210)),
     conf.int=FALSE, lwd=2, col="orange", lty="dashed")
lines(survfit(model.2,
     newdata=data.frame(rx=0, age=5, wbc=210)),
     conf.int=FALSE, lwd=2, col="orange", lty="solid")
legend("topright", cex=0.7,
       legend=c("treated, age=5, wbc=45", 
                "untreated, age=5, wbc=45",
                "treated, age=5, wbc=210",
                "untreated, age=5, wbc=210"),
       col=c("blue", "blue", "orange", "orange"),
       lty=c("dashed", "solid", "dashed", "solid"),
       lwd=c(2))

```



\newpage

####Question 2

```{r load_data}

link = "https://github.com/dmccoomes/Survival/raw/master/Homework%201/addicts.csv"
adix <- read.csv(link)

surv.adix <- Surv(time=adix$time, event=adix$event, type="right")
survfit.adix <- survfit(surv.adix ~ 1, data=adix, conf.type = "log-log")


```


(a) The HR of exit from maintenance comparing individuals who receive a one mg higher dose of methadone is 0.965 (95% CI: 0.953-0.977). The HR comparing those who were incarcerated to those who were not is 1.386 (95% CI: 0.999-1.924).


```{r}

model.2.1 <- coxph(surv.adix ~ dose + prison + clinic, data=adix)
summary(model.2.1)

```

