---
title: "Survival - Final Project"
author: "David Coomes"
date: "February 26, 2020"
output: 
   html_document:
     theme: cosmo
---


```{r install_knitr, include=FALSE}

if (!require(knitr)) install.packages("knitr")
library(knitr)

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r install_packages, include=FALSE}

if (!require(kableExtra)) install.packages("kableExtra")
if (!require(flexsurv)) install.packages("flexsurv")
if (!require(survMisc)) install.packages("survMisc")
if (!require(msm)) install.packages("msm")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(tableone)) install.packages("tableone")
if (!require(compareGroups)) install.packages("compareGroups")
if (!require(table1)) install.packages("table1")
if (!require(compareGroups)) install.packages("compareGroups", dep=TRUE)

```



```{r load_packages, include=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
#source("getmedianres.R")
library(survival) 
library(flexsurv)
library(survMisc) # for Wilcoxon-Gehan-Breslow test 
library(msm)
library(tableone)
#library(compareGroups) - can't install package "officer"
library(table1)


```


```{r load_data}

link = "https://github.com/dmccoomes/Survival/raw/master/Final%20Project/Data/bmt.csv"

data <- read.csv(url(link), header=TRUE)

```



```{r fit_survival}

surv.luke.1 <- Surv(time=data$tdfs, event=data$deltadfs, type="right")

```

<br>

## **Question 1**

<br>

```{r fit_parametric, include=FALSE}

#fit KM curve
survfit.luke.1 <- survfit(surv.luke.1 ~ 1, data=data, conf.type="log-log")
summary(survfit.luke.1)

```


```{r plot_km}

#plot KM curve
plot(survfit.luke.1,
     conf.int=TRUE,
     main="KM survival estimate of entire cohort (includes 95% CI)",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "light blue", "light blue"),
     lty = c("solid", "dashed", "dashed"),
     lwd = c(2,1,1))
     

```

There were 137 individuals in the cohort. During the study period, 83 (61%) of them had the event of interest (relapse or death). The median survival time in this cohort was 481 days (95% CI: 363-748). 

```{r, include=FALSE}

summary(survfit.luke.1)$table

```

<br>

## **Question 2:**

Generating a table 1.

```{r table_1, include=FALSE}

CreateTableOne(data=data)


```


<br>

## **Question 3:**
  
  We're interested in finding the association of any of the baseline factors with the outcome:
  
   - patient age                                                       (age: years)
   - patient gender                                                    (male: 1=male, 0=female)
   - donor age                                                         (donorage: years)
   - donor gender                                                      (donormale: 1=male, 0=female)
   - patient CMV status                                                (cmv: 1=CMV positive, 0=CMV negative)
   - donor CMV status                                                  (donorcmv: 1=CMV+, 0=CMV-)
   - wait time from diagnosis to transplant                            (waittime: waiting time until transplant (in days))
   - disease group                                                     (disgroup: 1=ALL, 2=AML low risk, 3=AML high risk)
   - FAB classification                                                (fab: 1=FAB grade 4 or 5 and AML, 0=otherwise)
   - prophylactic use of methotrexate to prevent aGVHD                 (mtx: 1=Yes, 0=No)

<br>


**Table 1: Baseline factor association with survival using a univariate model**

```{r multiple_univariate}

#run each cox ph model
covariates <- c("age", "male", "donorage", "donormale", "cmv", "donorcmv", "waittime", "disgroup", "fab", "mtx")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('surv.luke.1 ~', x)))
                        
univ_models <- lapply(univ_formulas, function(x){coxph(x, data=data)})

#get results
univ_results <- lapply(univ_models,
                       function(x){
                          x <- summary(x)
                          p.value <- round(x$wald["pvalue"], 3)
                          wald.test <- round(x$wald["test"], 3)
                          beta <- signif(x$coef[1], 3); 
                          HR <- round(x$coef[2], 3);
                          HR.confint.lower <- round(x$conf.int[,"lower .95"], 3)
                          HR.confint.upper <- round(x$conf.int[,"upper .95"], 3)
                          res <- c(HR, HR.confint.lower, HR.confint.upper, p.value)
                          names(res) <- c("HR", "95% CI lower", "95% CI upper", "p value")
                          return(res)
                       })

res <- t(as.data.frame(univ_results, check.names=FALSE))

#put results into a table
kable(res, caption="Cox PH HR using univariate model")  %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, position="left") %>%
   column_spec(1, width="4cm") %>%
   row_spec(9:10, bold=T)



```

<br>

**Table 2: Baseline factor association with survival while adjusting for other baseline factors**

```{r fit_baseline, include=FALSE}

#first, fit a Cox PH model including all baseline covariates
model.base.1 <- coxph(surv.luke.1 ~ age + male + donorage + donormale + cmv + donorcmv + waittime + disgroup + fab + mtx, data=data)
summary(model.base.1)

#trying a Cox PH model that includes hospital as a stratified variable
model.base.2 <- coxph(surv.luke.1 ~ age + male + donorage + donormale + cmv + donorcmv + waittime + disgroup + fab + mtx + strata(hospital), data=data)

#creating Cox PH model that includes country as a stratified variable
#creating country variable
data$country_us = ifelse((data$hospital==1 | data$hospital==4), 1, 0)
model.base.3 <- coxph(surv.luke.1 ~ age + male + donorage + donormale + cmv + donorcmv + waittime + disgroup + fab + mtx + strata(country_us), data=data)

#Creating model with just FAB and not including disease group
model.base.4 <- coxph(surv.luke.1 ~ age + male + donorage + donormale + cmv + donorcmv + waittime + fab + mtx, data=data)
summary(model.base.4)



#second, are any of these variables associated with difference by themselves - I did this in the box above
# model.base.age <- coxph(surv.luke.1 ~ age, data=data)
# model.base.male <- coxph(surv.luke.1 ~ male, data=data)
# model.base.donorage <- coxph(surv.luke.1 ~ donorage, data=data)
# model.base.donormale <- coxph(surv.luke.1 ~ donormale, data=data)
# model.base.cmv <- coxph(surv.luke.1 ~ cmv, data=data)
# model.base.donorcmv <- coxph(surv.luke.1 ~ donorcmv, data=data)
# model.base.waittime <- coxph(surv.luke.1 ~ waittime, data=data)
# model.base.disgroup <- coxph(surv.luke.1 ~ disgroup, data=data)
# model.base.fab <- coxph(surv.luke.1 ~ fab, data=data)
# model.base.mtx <- coxph(surv.luke.1 ~ mtx, data=data)
# 
# summary(model.base.age)
# summary(model.base.male)
# summary(model.base.donorage)
# summary(model.base.donormale)
# summary(model.base.cmv)
# summary(model.base.donorcmv)
# summary(model.base.waittime)
# summary(model.base.disgroup)
# summary(model.base.fab)
# summary(model.base.mtx)

#build table of z-scores for individual models

```


```{r model_summary, include=FALSE}

summary(model.base.1)

#mtx doesn't vary by hospital - only the hospitals in Australia used MTX as a prophylactic
#this means that we can't include MTX and hospital in the analysis - they are the same
summary(model.base.2)

summary(model.base.3)

#FAB is significant (0.0160) and MTX is significant (0.080). 

```



```{r}

p.value <- as.data.frame(summary(model.base.1)$coefficients[,5])
HR <- as.data.frame(summary(model.base.1)$coefficients[,2])
conf.int.low <- as.data.frame(summary(model.base.1)$conf.int[,3])
conf.int.high <- as.data.frame(summary(model.base.1)$conf.int[,4])

est <- cbind(HR, conf.int.low, conf.int.high, p.value)
names(est) <- c("HR", "HR 95% CI low", "HR 95% CI high", "p value")
est <- round(est, digits=3)

kable(est, caption="Cox PH HR using multivariate model")  %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, position="left") %>%
   column_spec(1, width="4cm") %>%
   row_spec(9:10, bold=T)

```





```{r logrank_test, include=FALSE}

survdiff(surv.luke.1 ~ male, data=data, rho=0)             #p-value 0.3
survdiff(surv.luke.1 ~ donormale, data=data, rho=0)        #        1
survdiff(surv.luke.1 ~ disgroup, data=data, rho=0)         #        0.001**
survdiff(surv.luke.1 ~ cmv, data=data, rho=0)              #        0.5
survdiff(surv.luke.1 ~ donorcmv, data=data, rho=0)         #        0.8
survdiff(surv.luke.1 ~ fab, data=data, rho=0)              #        0.004**
survdiff(surv.luke.1 ~ mtx, data=data, rho=0)              #        0.09*

```



```{r plot_one_NA_compare}

survfit.luke.disgroup <- survfit(surv.luke.1 ~ disgroup, data=data, conf.type="log-log")
plot(survfit.luke.disgroup,
     fun="cumhaz",
     col=c("blue", "red", "orange"), lwd=2, lty=c("solid", "solid", "dashed"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by disease group")
legend("topleft", c("ALL", "AML low", "AML high"), col=c("blue", "red", "orange"), lwd=c(2,2,2), lty=c("solid", "solid", "dashed"))


survfit.luke.fab <- survfit(surv.luke.1 ~ fab, data=data, conf.type="log-log")
plot(survfit.luke.fab,
     fun="cumhaz",
     col=c("blue", "red"), lwd=2, lty=c("solid", "solid"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by FAB")
legend("topleft", c("FAB 4 or 5 & AML", "other"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "solid"))


survfit.luke.mtx <- survfit(surv.luke.1 ~ mtx, data=data, conf.type="log-log")
plot(survfit.luke.mtx,
     fun="cumhaz",
     col=c("blue", "red"), lwd=2, lty=c("solid", "solid"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by MTX")
legend("bottomright", c("No methotrexate use", "Methotrexate use"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "solid"))

```


```{r}

#plot KM curve
plot(survfit.luke.disgroup,
     conf.int=FALSE,
     main="KM survival estimate by disease group",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "orange", "red"),
     lty = c("solid", "dashed", "solid"),
     lwd = c(2,2,2))
legend("topright", c("ALL", "AML low", "AML high"), col=c("blue", "orange", "red"), lwd=c(2,2,2), lty=c("solid", "dashed", "solid"))


plot(survfit.luke.fab,
     conf.int=FALSE,
     main="KM survival estimate by FAB classification",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "red"),
     lty = c("solid", "dashed"),
     lwd = c(2,2))
legend("topright", c("FAB 4 or 5 & AML", "other"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "dashed"))

plot(survfit.luke.mtx,
     conf.int=FALSE,
     main="KM survival estimate by Methotrexate use",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "red"),
     lty = c("solid", "dashed"),
     lwd = c(2,2))
legend("topright", c("No methotrexate use", "Methotrexate use"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "dashed"))

```


```{r wilcoxon_gehan_breslow_luke}

library(survMisc)

survfit.luke.male <- survfit(surv.luke.1 ~ male, data=data, conf.type="log-log")
survfit.luke.donormale <- survfit(surv.luke.1 ~ donormale, data=data, conf.type="log-log")
survfit.luke.cmv <- survfit(surv.luke.1 ~ cmv, data=data, conf.type="log-log")
survfit.luke.donorcmv <- survfit(surv.luke.1 ~ donorcmv, data=data, conf.type="log-log")
survfit.luke.mtx <- survfit(surv.luke.1 ~ mtx, data=data, conf.type="log-log")


comp(ten(survfit.luke.male))$tests          #p-value = 0.236
comp(ten(survfit.luke.donormale))$tests     #          0.930
comp(ten(survfit.luke.disgroup))$tests      #          0.051*
comp(ten(survfit.luke.cmv))$tests           #          0.397
comp(ten(survfit.luke.donorcmv))$tests      #          0.744
comp(ten(survfit.luke.fab))$tests           #          0.008**
comp(ten(survfit.luke.mtx))$tests           #          0.019**




```


<br>

Of the measured baseline variables, when using a Cox PH model, the FAB classification and prophylactic use of methotrexate (MTX) are significantly associated with differences in disease free survival. For FAB classification, this association is significant at the 90% confidence level using a univariate Cox PH model with FAB as the independent variable (p=0.004) and it remains significant after adjusting for all other baseline characteristics in a multivariate Cox PH model (p=0.016). For MTX, this association is significant at the 90% confidence level using a univariate Cox PH model with MTX as the independent variable (p=0.094) and it remains significant after adjusting for all other baseline characteristics in a multivariate Cox PH model (p=0.080).   

When fiting non-parametric survival models, and comparing Nelson-Aelen Cumulative Hazards, there are three baseline variables that are significantly associated with differences in disease free survival. Using the logrank test to compare groups, disease group (p=0.001), FAB classification (p=0.004), and MTX use (p=0.09) are all significantly associated with differences in cumulative hazard at the 90% confidence level. Using the Wilcoxan-Gehan-Breslow test, these three baseline measurements remain significantly different at the 90% confidence level. 




### Question 7 

Based on the available data, is recovery of normal platelet levels associated to improved disease-free survival? Is it associated with a decreased risk of relapse?

I believe this is a question that includes time varying covariates:
  There are two questions:
      1. Outcome = disease-free survival (same as above)
      Time-varying covariate:
         tp = time until recovery of normal platelet level (in days)
         deltap = recovery of normal platelet levels (0=no, 1=yes)
      2. Outcome = relapse
      Time-varying covariate:
         tp = time until recovery of normal platelet level (in days)
         deltap = recovery of normal platelet levels (0=no, 1=yes)
    
    
If we include an interaction in our model to account for age (of both donor and recipient), then which outcomes should we report?
     1. Quantiles of age?

```{r generating_dataset}

#setting up TVC dataset
luke.tvc = tmerge(data1=data,
                  data2=data,
                  id=id,
                  deltadfs=event(tdfs, deltadfs),
                  post_tp=tdc(tp))

#fitting TVC Cox PH model with no covariates
surv.luke.tvc <- Surv(luke.tvc$tstart, luke.tvc$tstop, luke.tvc$deltadfs)
model.tvc <- coxph(surv.luke.tvc ~ post_tp, data=luke.tvc)

summary(model.tvc)


#fitting TVC Cox PH model with covariates
model.tvc.cov <- coxph(surv.luke.tvc ~ post_tp + age + donorage + fab + male + donormale + waittime + cmv + donorcmv + mtx, data=luke.tvc)

summary(model.tvc.cov)

#with covariates by strata
luke.tvc$new_age <- luke.tvc$age - 28
luke.tvc$new_donorage <- luke.tvc$donorage - 28

model.tvc.cov.strata <- coxph(surv.luke.tvc ~ post_tp + age + donorage*post_tp + male + donormale + waittime + strata(cmv) + strata(mtx)  + strata(donorcmv) + strata(fab), data=luke.tvc)

summary(model.tvc.cov.strata)





#I think this is the model I want to use
model.tvc.cov.strata.2 <- coxph(surv.luke.tvc ~ post_tp + new_age*post_tp + new_donorage*post_tp + male + donormale + waittime + strata(cmv) + strata(mtx)  + strata(donorcmv) + strata(fab) + fab*post_tp, data=luke.tvc)

summary(model.tvc.cov.strata.2)






luke.tvc$fab_tp <- luke.tvc$fab*luke.tvc$post_tp
luke.tvc$age_tp <- luke.tvc$age*luke.tvc$post_tp
luke.tvc$donorage_tp <- luke.tvc$donorage*luke.tvc$post_tp
luke.tvc$male_tp <- luke.tvc$male*luke.tvc$post_tp
luke.tvc$donormale_tp <- luke.tvc$donormale*luke.tvc$post_tp
luke.tvc$cmv_tp <- luke.tvc$cmv*luke.tvc$post_tp
luke.tvc$donorcmv_tp <- luke.tvc$donorcmv*luke.tvc$post_tp
luke.tvc$mtx_tp <- luke.tvc$mtx*luke.tvc$post_tp
luke.tvc$waittime_tp <- luke.tvc$waittime*luke.tvc$post_tp

#with interaction terms
# model.tvc.int <- coxph(surv.luke.tvc ~ post_tp + fab_tp + age_tp + donorage_tp + male_tp + donormale_tp + cmv_tp + donorcmv_tp +mtx_tp + waittime_tp, data=luke.tvc)
# 
# summary(model.tvc.int)

#better one with interaction terms
model.tvc.int.2 <- coxph(surv.luke.tvc ~ post_tp + age*post_tp + donorage*post_tp + male*post_tp + donormale*post_tp + waittime*post_tp + cmv*post_tp + donorcmv*post_tp + mtx*post_tp + fab*post_tp, data=luke.tvc)

summary(model.tvc.int.2)

#stratified and interactions
model.tvc.int.3 <- coxph(surv.luke.tvc ~ post_tp + age*post_tp + donorage*post_tp + male + donormale + waittime + cmv + donorcmv + mtx + fab, data=luke.tvc)

summary(model.tvc.int.3)

```


```{r age_interaction}

#we would have a different value for each age - I don't think that's worth reporting
# est.age <- exp(coef(model.tvc.int.3)["post_tp"] + 
#                  coef(model.tvc.int.3)["age"] +
#                  coef(model.tvc.int.3)["post_tp:age"])
# 
# se.20 <- deltamethod(g=~exp(x1 + x2 + x15))
# 

```


Based on the available data, without adjusting for any covariates, recovery of normal platelet count is associated with a 72.1% (95% CI: 46.9-85.3%) lower hazard of relapse or death as compared to those whose platelet count did not return to normal. 

After adjusting for multiple covariates, recovery of normal platelet count is associated with a 70.2% (95% CI: 41.5-84.8%) lower hazard of relapse or death as compared to those whose platelet count did not return to normal. 

If we adjust for cmv, donorcmv, and mtx by stratified PH models, then recovery of normal platelet count is associated with a 65.9% (95% CI: 24.7-84.5%) lower hazard of relapse or death as compared to those whose platelet count did not return to normal. 

If we include age and donorage as interaction terms within our model, 


```{r age_quantiles}

library(gtools)
luke.tvc$age_cat <- cut(luke.tvc$age, seq(0, 60, 10)) 
luke.tvc$donorage_cat <- cut(luke.tvc$donorage, seq(0, 60, 10)) 

#combining lowest and highest categories with cat next to them
luke.tvc$age_cat <- cut(luke.tvc$age, c(0, 20, 30, 40, 60)) 
luke.tvc$donorage_cat <- cut(luke.tvc$donorage, c(0, 20, 30, 40, 60)) 
luke.tvc$age_bin <- as.numeric(luke.tvc$age_cat)
luke.tvc$donorage_bin <- as.numeric(luke.tvc$donorage_cat)


model.tvc.int.4 <- coxph(surv.luke.tvc ~ post_tp + age_cat*post_tp + donorage_cat*post_tp + male + donormale + waittime + cmv + donorcmv + mtx + fab, data=luke.tvc)

summary(model.tvc.int.4)

#age as a grouped linear predictor
model.tvc.int.5 <- coxph(surv.luke.tvc ~ post_tp + age_bin*post_tp + donorage_bin*post_tp + male + donormale + waittime + cmv + donorcmv + mtx + fab, data=luke.tvc)

summary(model.tvc.int.5)

#creating a binary age factor - this doesn't seem to be significant
luke.tvc$age_old <- ifelse(luke.tvc$age >28, 1, 0)
luke.tvc$donorage_old <- ifelse(luke.tvc$donorage >28, 1, 0)

model.tvc.int.6 <- coxph(surv.luke.tvc ~ post_tp + age*post_tp + donorage*post_tp + fab + male + donormale + waittime + cmv + donorcmv + mtx, data=luke.tvc)

summary(model.tvc.int.6)


```


```{r}

est.20 <- exp(coef(model.tvc.int.5)["post_tp"] + 
                 coef(model.tvc.int.5)["post_tp:age_bin"])

se.20 <- deltamethod(g=~exp((x1 + x2)),
                     mean=coef(model.tvc.int.5)[c("post_tp", "post_tp:age_bin")],
                     cov=vcov(model.tvc.int.5)[c("post_tp", "post_tp:age_bin"),
                                               c("post_tp", "post_tp:age_bin")],
                     ses=TRUE)

#age group 20-30
as.numeric(est.20)
as.numeric(c(est.20 - 1.96*se.20,
             est.20 + 1.96*se.20))


est.30 <- exp(coef(model.tvc.int.5)["post_tp"] + 
                 2*coef(model.tvc.int.5)["post_tp:age_bin"])

se.30 <- deltamethod(g=~exp((x1 + 2*x2)),
                     mean=coef(model.tvc.int.5)[c("post_tp", "post_tp:age_bin")],
                     cov=vcov(model.tvc.int.5)[c("post_tp", "post_tp:age_bin"),
                                               c("post_tp", "post_tp:age_bin")],
                     ses=TRUE)

#age group 30-40
as.numeric(est.30)
as.numeric(c(est.30 - 1.96*se.30,
             est.30 + 1.96*se.30))


est.40 <- exp(coef(model.tvc.int.5)["post_tp"] + 
                 3*coef(model.tvc.int.5)["post_tp:age_bin"])

se.40 <- deltamethod(g=~exp((x1 + 3*x2)),
                     mean=coef(model.tvc.int.5)[c("post_tp", "post_tp:age_bin")],
                     cov=vcov(model.tvc.int.5)[c("post_tp", "post_tp:age_bin"),
                                               c("post_tp", "post_tp:age_bin")],
                     ses=TRUE)

#age group 40-50
as.numeric(est.40)
as.numeric(c(est.40 - 1.96*se.40,
             est.40 + 1.96*se.40))


est.50 <- exp(coef(model.tvc.int.5)["post_tp"] + 
                 4*coef(model.tvc.int.5)["post_tp:age_bin"])

se.50 <- deltamethod(g=~exp((x1 + 4*x2)),
                     mean=coef(model.tvc.int.5)[c("post_tp", "post_tp:age_bin")],
                     cov=vcov(model.tvc.int.5)[c("post_tp", "post_tp:age_bin"),
                                               c("post_tp", "post_tp:age_bin")],
                     ses=TRUE)

#age group 20-30
as.numeric(est.50)
as.numeric(c(est.50 - 1.96*se.50,
             est.50 + 1.96*se.50))





```




```{r compare_groups}

#looking at the length of time for platelet recovery
#those that did not recover
summary(subset(luke.tvc, post_tp==0, select=c(tstart, tstop)))
#those that did recover
summary(subset(luke.tvc, post_tp==1, select=c(tstart, tstop)))

```



