---
title: "Survival - Final Project"
author: "David Coomes"
date: "February 26, 2020"
output: 
   html_document:
     theme: cosmo
---


```{r install_knitr, include=FALSE}

if (!require(knitr)) install.packages("knitr")
library(knitr)

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r install_packages, include=FALSE}

if (!require(kableExtra)) install.packages("kableExtra")
if (!require(flexsurv)) install.packages("flexsurv")
if (!require(survMisc)) install.packages("survMisc")
if (!require(msm)) install.packages("msm")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(tableone)) install.packages("tableone")
if (!require(compareGroups)) install.packages("compareGroups")
if (!require(table1)) install.packages("table1")

```



```{r load_packages, include=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
#source("getmedianres.R")
library(survival) 
library(flexsurv)
library(survMisc) # for Wilcoxon-Gehan-Breslow test 
library(msm)
library(tableone)
library(compareGroups)
library(table1)


```


```{r load_data}

link = "https://github.com/dmccoomes/Survival/raw/master/Final%20Project/Data/bmt.csv"

data <- read.csv(url(link), header=TRUE)

```



```{r fit_survival}

surv.luke.1 <- Surv(time=data$tdfs, event=data$deltadfs, type="right")

```

<br>

## **Question 1**

<br>

```{r fit_parametric, include=FALSE}

#fit KM curve
survfit.luke.1 <- survfit(surv.luke.1 ~ 1, data=data, conf.type="log-log")
summary(survfit.luke.1)

```


```{r plot_km}

#plot KM curve
plot(survfit.luke.1,
     conf.int=TRUE,
     main="KM survival estimate of entire cohort (includes 95% CI)",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "light blue", "light blue"),
     lty = c("solid", "dashed", "dashed"),
     lwd = c(2,1,1))
     

```

There were 137 individuals in the cohort. During the study period, 83 (61%) of them had the event of interest (relapse or death). The median survival time in this cohort was 481 days (95% CI: 363-748). 

```{r, include=FALSE}

summary(survfit.luke.1)$table

```

<br>

## **Question 2:**

Generating a table 1.

```{r}

CreateTableOne(data=data)


```


<br>

## **Question 3:**
  
  We're interested in finding the association of any of the baseline factors with the outcome:
  
   - patient age                                                       (age: years)
   - patient gender                                                    (male: 1=male, 0=female)
   - donor age                                                         (donorage: years)
   - donor gender                                                      (donormale: 1=male, 0=female)
   - patient CMV status                                                (cmv: 1=CMV positive, 0=CMV negative)
   - donor CMV status                                                  (donorcmv: 1=CMV+, 0=CMV-)
   - wait time from diagnosis to transplant                            (waittime: waiting time until transplant (in days))
   - disease group                                                     (disgroup: 1=ALL, 2=AML low risk, 3=AML high risk)
   - FAB classification                                                (fab: 1=FAB grade 4 or 5 and AML, 0=otherwise)
   - prophylactic use of methotrexate to prevent aGVHD                 (mtx: 1=Yes, 0=No)

<br>


**Table 1: Baseline factor association with survival using a univariate model**

```{r multiple_univariate}

#run each cox ph model
covariates <- c("age", "male", "donorage", "donormale", "cmv", "donorcmv", "waittime", "disgroup", "fab", "mtx")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('surv.luke.1 ~', x)))
                        
univ_models <- lapply(univ_formulas, function(x){coxph(x, data=data)})

#get results
univ_results <- lapply(univ_models,
                       function(x){
                          x <- summary(x)
                          #p.value <- signif(x$wald["pvalue"], digits=3)
                          p.value <- round(x$wald["pvalue"], 3)
                          wald.test <- round(x$wald["test"], 3)
                          beta <- signif(x$coef[1], 3); 
                          HR <- round(x$coef[2], 3);
                          HR.confint.lower <- round(x$conf.int[,"lower .95"], 3)
                          HR.confint.upper <- round(x$conf.int[,"upper .95"], 3)
                          #HR <- paste0(HR, " (",
                          #             HR.confint.lower, "-", HR.confint.upper, ")")
                          #res <- c(beta, HR, wald.test, p.value)       only including HR, 95% CI, and p value
                          res <- c(HR, HR.confint.lower, HR.confint.upper, p.value)
                          #names(res) <- c("beta", "HR (95% CI for HR", "wald.test", "p.value")
                          names(res) <- c("HR", "95% CI lower", "95% CI upper", "p value")
                          return(res)
                       })

res <- t(as.data.frame(univ_results, check.names=FALSE))
#as.data.frame(res)


kable(res, caption="Cox PH HR using univariate model")  %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, position="left") %>%
   column_spec(1, width="4cm") %>%
   row_spec(9:10, bold=T)



```

<br>

**Table 2: Baseline factor association with survival while adjusting for other baseline factors**

```{r fit_baseline, include=FALSE}

#first, fit a Cox PH model including all baseline covariates

model.base.1 <- coxph(surv.luke.1 ~ age + male + donorage + donormale + cmv + donorcmv + waittime + disgroup + fab + mtx, data=data)
summary(model.base.1)

#second, are any of these variables associated with difference by themselves - I did this in the box above
model.base.age <- coxph(surv.luke.1 ~ age, data=data)
model.base.male <- coxph(surv.luke.1 ~ male, data=data)
model.base.donorage <- coxph(surv.luke.1 ~ donorage, data=data)
model.base.donormale <- coxph(surv.luke.1 ~ donormale, data=data)
model.base.cmv <- coxph(surv.luke.1 ~ cmv, data=data)
model.base.donorcmv <- coxph(surv.luke.1 ~ donorcmv, data=data)
model.base.waittime <- coxph(surv.luke.1 ~ waittime, data=data)
model.base.disgroup <- coxph(surv.luke.1 ~ disgroup, data=data)
model.base.fab <- coxph(surv.luke.1 ~ fab, data=data)
model.base.mtx <- coxph(surv.luke.1 ~ mtx, data=data)

summary(model.base.age)
summary(model.base.male)
summary(model.base.donorage)
summary(model.base.donormale)
summary(model.base.cmv)
summary(model.base.donorcmv)
summary(model.base.waittime)
summary(model.base.disgroup)
summary(model.base.fab)
summary(model.base.mtx)

#build table of z-scores for individual models

```


```{r}

summary(model.base.1)

#FAB is significant (0.0160) and MTX is significant (0.080). 

```



```{r}

p.value <- as.data.frame(summary(model.base.1)$coefficients[,5])
HR <- as.data.frame(summary(model.base.1)$coefficients[,2])
conf.int.low <- as.data.frame(summary(model.base.1)$conf.int[,3])
conf.int.high <- as.data.frame(summary(model.base.1)$conf.int[,4])

est <- cbind(HR, conf.int.low, conf.int.high, p.value)
names(est) <- c("HR", "HR 95% CI low", "HR 95% CI high", "p value")
est <- round(est, digits=3)
#est

kable(est, caption="Cox PH HR using multivariate model")  %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, position="left") %>%
   column_spec(1, width="4cm") %>%
   row_spec(9:10, bold=T)

```



```{r logrank_test, include=FALSE}

survdiff(surv.luke.1 ~ male, data=data, rho=0)             #p-value 0.3
survdiff(surv.luke.1 ~ donormale, data=data, rho=0)        #        1
survdiff(surv.luke.1 ~ disgroup, data=data, rho=0)         #        0.001**
survdiff(surv.luke.1 ~ cmv, data=data, rho=0)              #        0.5
survdiff(surv.luke.1 ~ donorcmv, data=data, rho=0)         #        0.8
survdiff(surv.luke.1 ~ fab, data=data, rho=0)              #        0.004**
survdiff(surv.luke.1 ~ mtx, data=data, rho=0)              #        0.09*

```



```{r plot_one_NA_compare}

survfit.luke.disgroup <- survfit(surv.luke.1 ~ disgroup, data=data, conf.type="log-log")
plot(survfit.luke.disgroup,
     fun="cumhaz",
     col=c("blue", "red", "orange"), lwd=2, lty=c("solid", "solid", "dashed"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by disease group")
legend("topleft", c("ALL", "AML low", "AML high"), col=c("blue", "red", "orange"), lwd=c(2,2,2), lty=c("solid", "solid", "dashed"))


survfit.luke.fab <- survfit(surv.luke.1 ~ fab, data=data, conf.type="log-log")
plot(survfit.luke.fab,
     fun="cumhaz",
     col=c("blue", "red"), lwd=2, lty=c("solid", "solid"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by FAB")
legend("topleft", c("FAB 4 or 5 & AML", "other"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "solid"))


survfit.luke.mtx <- survfit(surv.luke.1 ~ mtx, data=data, conf.type="log-log")
plot(survfit.luke.mtx,
     fun="cumhaz",
     col=c("blue", "red"), lwd=2, lty=c("solid", "solid"),
     xlab="Time (in days)", ylab="Cumulative Hazard", main="Nelson-Aalen cumulative hazard estimates by MTX")
legend("bottomright", c("No methotrexate use", "Methotrexate use"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "solid"))

```


```{r}

#plot KM curve
plot(survfit.luke.disgroup,
     conf.int=FALSE,
     main="KM survival estimate by disease group",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "orange", "red"),
     lty = c("solid", "dashed", "solid"),
     lwd = c(2,2,2))
legend("topright", c("ALL", "AML low", "AML high"), col=c("blue", "red", "orange"), lwd=c(2,2,2), lty=c("solid", "solid", "dashed"))


plot(survfit.luke.fab,
     conf.int=FALSE,
     main="KM survival estimate by FAB classification",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "red"),
     lty = c("solid", "dashed"),
     lwd = c(2,2))
legend("topright", c("FAB 4 or 5 & AML", "other"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "dashed"))

plot(survfit.luke.mtx,
     conf.int=FALSE,
     main="KM survival estimate by Methotrexate use",
     ylab = "Survival probability",
     xlab = "Time (days)",
     col = c("blue", "red"),
     lty = c("solid", "dashed"),
     lwd = c(2,2))
legend("topright", c("No methotrexate use", "Methotrexate use"), col=c("blue", "red"), lwd=c(2,2), lty=c("solid", "dashed"))

```


```{r wilcoxon_gehan_breslow_luke}

library(survMisc)

survfit.luke.male <- survfit(surv.luke.1 ~ male, data=data, conf.type="log-log")
survfit.luke.donormale <- survfit(surv.luke.1 ~ donormale, data=data, conf.type="log-log")
survfit.luke.cmv <- survfit(surv.luke.1 ~ cmv, data=data, conf.type="log-log")
survfit.luke.donorcmv <- survfit(surv.luke.1 ~ donorcmv, data=data, conf.type="log-log")
survfit.luke.mtx <- survfit(surv.luke.1 ~ mtx, data=data, conf.type="log-log")


comp(ten(survfit.luke.male))$tests          #p-value = 0.236
comp(ten(survfit.luke.donormale))$tests     #          0.930
comp(ten(survfit.luke.disgroup))$tests      #          0.051*
comp(ten(survfit.luke.cmv))$tests           #          0.397
comp(ten(survfit.luke.donorcmv))$tests      #          0.744
comp(ten(survfit.luke.fab))$tests           #          0.008**
comp(ten(survfit.luke.mtx))$tests           #          0.019**




```


<br>

Of the measured baseline variables, when using a Cox PH model, the FAB classification and prophylactic use of methotrexate (MTX) are significantly associated with differences in disease free survival. For FAB classification, this association is significant at the 90% confidence level using a univariate Cox PH model with FAB as the independent variable (p=0.004) and it remains significant after adjusting for all other baseline characteristics in a multivariate Cox PH model (p=0.016). For MTX, this association is significant at the 90% confidence level using a univariate Cox PH model with MTX as the independent variable (p=0.094) and it remains significant after adjusting for all other baseline characteristics in a multivariate Cox PH model (p=0.080).   

When fiting non-parametric survival models, and comparing Nelson-Aelen Cumulative Hazards, there are three baseline variables that are significantly associated with differences in disease free survival. Using the logrank test to compare groups, disease group (p=0.001), FAB classification (p=0.004), and MTX use (p=0.09) are all significantly associated with differences in cumulative hazard at the 90% confidence level. Using the Wilcoxan-Gehan-Breslow test, these three baseline measurements remain significantly different at the 90% confidence level. 
