---
title: "EPI/BIOST 537 - Discussion 4"
author: "Eric Morenz (Prepared by Jeremy Roth)"
date: 'February 4, 2019'
output:
  html_document: default
  pdf_document: default
---

## Load in the herpes dataset and create the apporopriate survival/survfit objects
```{r load_data, message=FALSE, warning=FALSE}
library(survival) 
library(flexsurv)
library(survMisc) # for Wilcoxon-Gehan-Breslow test 
source("getmedianres.R")
herpes <- read.csv(file="Data/herpes.csv", header=TRUE)
surv.herpes <- Surv(time=herpes$timetorec, event=herpes$event, type="right")
survfit.herpes <- survfit(surv.herpes ~ 1, data=herpes, conf.type = "log-log")
```

## Problem 2
### 2(a) Plot the Kaplan-Meier estimator along with pointwise 95% CIs. What is the estimated probability that no event will occur by 120 days?
```{r plot_KM, fig.align="center"}
plot(survfit.herpes, main="Kaplan-Meier survival estimate", 
      ylab="Survival probability", xlab="Days")
abline(v=120, col="blue", lty="dashed")
axis(1, at=120, labels=120, col.axis='blue')
```

```{r}
summary(survfit.herpes, times=c(119, 120, 121))
```

### 2(b) Provide the estimated median survival time and associated 95% confidence interval by:
#### (i) scrutinizing values of the Kaplan-Meier estimator and associated confidence intervals
```{r}
summary(survfit.herpes, times=seq(1:70))
```

#### (ii) using the median estimate and confidence intervals provided by the survfit command
```{r}
summary(survfit.herpes)$table
```

### 2(c) Investigate differences between patients with and without a history of incarceration.
#### (i) On the same graph, plot the Kaplan-Meier estimates for male=0 and male=1
```{r, fig.align="center"}
survfit.herpes.by.male <- survfit(surv.herpes ~ male, data=herpes, conf.type = "log-log")
plot(survfit.herpes.by.male, 
       col=c("blue", "orange"), lwd=2, lty=c("solid", "dashed"), 
      xlab="Days", ylab="Survival probability", main="Kaplan-Meier survival estimates")
legend("topright", c("male=0", "male=1"), col=c("blue", "orange"), lwd=c(3, 3), lty=c("solid", "dashed"))
```

#### (ii) Does the probability that no event occurred by 8 months differ significantly between these two groups?
```{r}
# estimate and SE of S(8 months) in male==0 subset
herpes.male.0 <- herpes[herpes$male == 0, ]
surv.herpes.male.0 <- Surv(time=herpes.male.0$timetorec, event=herpes.male.0$event)
survfit.herpes.male.0 <- survfit(surv.herpes.male.0 ~ 1, data=herpes.male.0, conf.type = "log-log")
summary(survfit.herpes.male.0, times=240)

# estimate and SE of S(8 months) in male==1 subset
herpes.male.1 <- herpes[herpes$male == 1, ]
surv.herpes.male.1 <- Surv(time=herpes.male.1$timetorec, event=herpes.male.1$event)
survfit.herpes.male.1 <- survfit(surv.herpes.male.1 ~ 1, data=herpes.male.1, conf.type = "log-log")
summary(survfit.herpes.male.1, times=240)
# compute Wald test statistic
wald.statistic.male.8.months <- (summary(survfit.herpes.male.0, times=240)["surv"]$surv - 
                             summary(survfit.herpes.male.1, times=240)["surv"]$surv) / 
                             (summary(survfit.herpes.male.0, times=240)["std.err"]$std.err^2 + 
                             summary(survfit.herpes.male.1, times=240)["std.err"]$std.err^2)^(1/2)
wald.statistic.male.8.months
# compute Wald p-value
2 * pnorm(-abs(wald.statistic.male.8.months))
```

#### (iii) Based on the logrank test, does the distribution of survival time differ significantly between the two groups?
```{r}
survdiff(surv.herpes ~ male, data=herpes, rho=0) 
```


#### (iv) Based on the Wilcoxon-Gehan-Breslow test, does the distribution of survival time differ significantly between the two groups?
```{r}
# looking at the "n" row and the "pNorm" column
comp(ten(survfit.herpes.by.male))$tests
```
#### (v) Plot estimated hazard functions for patients with and without a history of incarceration.
```{r, fig.align="center"}
plot(survfit.herpes.by.male, 
     fun="cumhaz",
      col=c("blue", "orange"), lwd=2, lty=c("solid", "dashed"), 
      xlab="Days", ylab="Cumulative Hazard", main="Nelson−Aalen cumulative hazard estimates")
legend("topleft", c("male=0", "male=1"), col=c("blue", "orange"), lwd=c(3, 3), lty=c("solid", "dashed"))
```

### 2(e) Based on a stratified logrank test, does the distribution of survival time differs by sex, adjusting for duration of primary episode > 15 days? State explicitly what the null and alternative hypotheses are and contrast with what they are in a standard logrank test.
```{r}
herpes$duration.above.15 <- herpes$duration > 15
survdiff(surv.herpes ~ male + strata(duration.above.15), data=herpes)
```
+ In a standard log-rank test, the null hypothesis is that 
$$S_{\text{male}=0}(t) = S_{\text{male}=1}(t) $$
for all $t$.
+ In a stratified log-rank test, the null hypothesis is that 
$$ S_{\text{male}=0, \: \text{duration} \leq 15}(t) = S_{\text{male}=1, \: \text{duration} \leq 15}(t)$$
**and**
$$ S_{\text{male}=0, \: \text{duration} > 15}(t) = S_{\text{male}=1, \: \text{duration} > 15}(t)$$
for all $t$.

### 2(f). What is the estimated median residual survival time at 60 days?
#### (i) Calculate these estimates using only values of the Kaplan-Meier estimator.
```{r}
# need estimated survival probability at 60 days
summary(survfit.herpes, times=60)
# find half of the estimated survival probability at 60 days
0.5 * summary(survfit.herpes, times=60)["surv"]$surv
# find the t where S(t) goes below  0.2358444 for the first time
summary(survfit.herpes, times=seq(1:200))
# don't forget to subtract 60 days from this time
182 - 60
```

#### (ii) Verify your answer using the R function provided for this purpose, and obtain 95% confidence intervals to accompany your estimates.
```{r}
df.for.medresidual <- data.frame("y"=herpes$timetorec, "delta"=herpes$event)
medresidual.60 <- getmedianres(survobj=surv.herpes,
                              times=60, confint=TRUE)
## estimate of median residual survival time at 60 days
medresidual.60$estimates
## CI for median residual survival time at 60 days
c(medresidual.60$ci.lower, medresidual.60$ci.upper)
```

