---
title: "Survival HW4"
author: "David Coomes"
date: "Due March 9, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r install_packages, include=FALSE}

if (!require(kableExtra)) install.packages("kableExtra")
if (!require(flexsurv)) install.packages("flexsurv")
if (!require(survMisc)) install.packages("survMisc")
if (!require(msm)) install.packages("msm")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(tableone)) install.packages("tableone")
if (!require(compareGroups)) install.packages("compareGroups")
if (!require(table1)) install.packages("table1")

```



```{r load_packages, include=FALSE}

library(tidyverse)
library(knitr)
library(kableExtra)
#source("getmedianres.R")
library(survival) 
library(flexsurv)
library(survMisc) # for Wilcoxon-Gehan-Breslow test 
library(msm)
library(tableone)
library(compareGroups)
library(table1)


```


```{r load_data2}

link = "https://github.com/dmccoomes/Survival/raw/master/Homework%201/addicts.csv"
adix <- read.csv(link)

surv.adix <- Surv(time=adix$time, event=adix$event, type="right")
survfit.adix <- survfit(surv.adix ~ 1, data=adix, conf.type = "log-log")


```

### Question 1

(a) 

**Table 1. Time ratios for Accelerated Failure Time model**

```{r fit_weibull_AFT, echo=FALSE}

weibull.adix.1 <- flexsurvreg(surv.adix ~ dose + clinic + prison, data=adix, dist="weibull")
#weibull.adix.1

weibull.adix.est <- round(exp(weibull.adix.1$res[3:5, 1:3]), 3)

weibull.adix.est %>%
  kable(col.names = c("Estimate", "95% CI Lower", "95% CI Upper")) %>%
  kable_styling(full_width = F,
                position="left")

```
 <br>
 
My results from the AFT model agree qualitatively with my results from the Cox PH model. In the AFT model we see that there is a slight (and significant) increase in time to exit from maintenance associated with an increase in methadone dose, which corresponds to a slight decrease in HR in the Cox model, adjusting for clinic and history of previous incarceration for both models. Similarily, we see a large increase in time to exit associated with clinic 2 as compared to clinic 1, and a large reduction in HR. Finally, there is a small (and non-significant) decrease in time to exit from maintenance associated with a history of previous incarceration, which corresponds to an increase in the HR (also non-significant). 


When using a generalized gamma baseline distribution, the dose time ratio estimate decreases by 0.68%, the clinic time ratio increases by 0.79%, and the incarceration time ratio increases by 4.91%. 

```{r fit_coxPH, include=FALSE}

cox.1 <- coxph(surv.adix ~ dose + clinic + prison, data=adix)
cox.1

```


```{r AFT_gengamma, include=FALSE}

gengamma.adix.1 <- flexsurvreg(surv.adix ~ dose + clinic + prison, data=adix, dist="gengamma")
#gengamma.adix.1

gengamma.adix.est <- round(exp(gengamma.adix.1$res[4:6, 1:3]), 3)
#gengamma.adix.est

perc_diff <- (gengamma.adix.est[,1]-weibull.adix.est[,1])/weibull.adix.est[,1]

perc_diff

```

<br> 

(b) The ratio of mean time to exit comparing an individual from clinic 2 without a history of incarceration and administered a dose of 40 mg/day to individuals from clinic 1 with a history of incarceration and administered a dose of 100 mg/day is **0.590 (95% CI: 0.305-1.143).**




```{r AFT_compare, include=FALSE}

AFT.point <- (40-100) * coef(weibull.adix.1)["dose"] + 
                  (2-1) * coef(weibull.adix.1)["clinic"] + 
                  (0-1) * coef(weibull.adix.1)["prison"]

AFT.se <- deltamethod(g = ~(40-100) * x1 + (2-1) * x2 + (0-1) * x3,
                      mean = coef(weibull.adix.1)[c("dose", "clinic", "prison")],
                      cov=vcov(weibull.adix.1)[c("dose", "clinic", "prison"),
                                               c("dose", "clinic", "prison")],
                      ses=TRUE)

#point estimate
exp(AFT.point)

#confindene interval
exp(c(AFT.point - 1.96*AFT.se,
      AFT.point + 1.96*AFT.se))

```



The median time until exit for a person from clinic 2 without a history of incarceration who is administered a dose of 40 mg/day is **509 days (95% CI: 334-684).** The median time unitl exit for a person from clinic 1 with a history of incarceration who is administered a dose of 100 mg/day is **862 days (95% CI: 511-1213).**

```{r median_survival, include=FALSE}

#median for person from clinic 2 on 40 mg/day with no history of incarceration
AFT.median.1 <- exp(40 * coef(weibull.adix.1)["dose"] + 
                      2 * coef(weibull.adix.1)["clinic"]) * 
        log(2) ^ (1/exp(coef(weibull.adix.1)["shape"])) * 
        exp(coef(weibull.adix.1)["scale"])

AFT.median.1.se <- deltamethod(g = ~exp(40*x1 + 2*x2) 
                               * log(2) ^ (1/exp(x4)) * exp(x5),
                               mean = coef(weibull.adix.1)[c("dose", "clinic", "prison", "shape", "scale")],
                               cov=vcov(weibull.adix.1)[c("dose", "clinic", "prison", "shape", "scale"),
                                                        c("dose", "clinic", "prison", "shape", "scale")],
                               ses=TRUE)

#point estimate
AFT.median.1
#confidence interval
c(AFT.median.1 - 1.96*AFT.median.1.se,
  AFT.median.1 + 1.96*AFT.median.1.se)


#median for person from clinic 1 on 100 mg/day with a history of incarceration
AFT.median.2 <- exp(100 * coef(weibull.adix.1)["dose"] + 
                      1 * coef(weibull.adix.1)["clinic"] + 
                      1 * coef(weibull.adix.1)["prison"]) * 
        log(2) ^ (1/exp(coef(weibull.adix.1)["shape"])) * 
        exp(coef(weibull.adix.1)["scale"])

AFT.median.2.se <- deltamethod(g = ~exp(100*x1 + 1*x2 + 1*x3) 
                               * log(2) ^ (1/exp(x4)) * exp(x5),
                               mean = coef(weibull.adix.1)[c("dose", "clinic", "prison", "shape", "scale")],
                               cov=vcov(weibull.adix.1)[c("dose", "clinic", "prison", "shape", "scale"),
                                                        c("dose", "clinic", "prison", "shape", "scale")],
                               ses=TRUE)

#point estimate
AFT.median.2
#confidence interval
c(AFT.median.2 - 1.96*AFT.median.2.se,
  AFT.median.2 + 1.96*AFT.median.2.se)


```


<br>

(c) The time ratio comparing patients who were administered a dose of 80 mg/day to those administered 60 mg/day *with* a history of previous incarceration and from the same clinic is **1.507 (95% CI: 1.135-2.002).** The time ratio comparing patients who were administed a dose of 80 mg/day to those administered 60 mg/day *wihout* a history of previous incarceration and from the same clinic is **1.712 (95% CI: 1.362-2.152).** These two subgroup-specific time ratios are not significantly different from each other **(p=0.46)**.

```{r new_weibull, include=FALSE}

#fit new weibull AFT model
weibull.adix.2 <- flexsurvreg(surv.adix ~ dose + clinic + prison + dose*prison, data=adix, dist="weibull")

#weibull.adix.3 <- flexsurvreg(surv.adix ~ dose*prison + clinic, data=adix, dist="weibull")

#get point estimate and CI comparing 80 to 60 mg/day for someone with a history of incarceration and from the same clinic
AFT.point.2 <- (80-60) * coef(weibull.adix.2)["dose"] + 
                  (80-60) * coef(weibull.adix.2)["dose:prison"]

AFT.se.2 <- deltamethod(g = ~(80-60) * x1 + (80-60) * x4,
                        mean = coef(weibull.adix.2)[c("dose", "clinic", "prison", "dose:prison")],
                        cov=vcov(weibull.adix.2)[c("dose", "clinic", "prison", "dose:prison"),
                                                 c("dose", "clinic", "prison", "dose:prison")],
                        ses=TRUE)

#point estimate
exp(AFT.point.2)

#confidence interval
exp(c(AFT.point.2 - 1.96*AFT.se.2,
       AFT.point.2 + 1.96*AFT.se.2))


#get point estimate and CI comparing 80 to 60 mg/day for someone without a history of incarceration and from the same clinic
AFT.point.3 <- (80-60) * coef(weibull.adix.2)["dose"] 

AFT.se.3 <- deltamethod(g = ~(80-60) * x1,
                        mean = coef(weibull.adix.2)[c("dose", "clinic", "prison", "dose:prison")],
                        cov=vcov(weibull.adix.2)[c("dose", "clinic", "prison", "dose:prison"),
                                                 c("dose", "clinic", "prison", "dose:prison")],
                        ses=TRUE)

#point estimate
exp(AFT.point.3)

#confidence interval
exp(c(AFT.point.3 - 1.96*AFT.se.3,
       AFT.point.3 + 1.96*AFT.se.3))

```



```{r interaction_test, include=FALSE}

#using wald test
SE.int.term <- deltamethod(g = ~exp((80-60) * x1), 
                            mean=coef(weibull.adix.2)["dose:prison"],
                            cov=vcov(weibull.adix.2)["dose:prison", "dose:prison"],
                            ses=TRUE)
                           
wald.test.stat <- (exp((80-60) * coef(weibull.adix.2)["dose:prison"])-1) / SE.int.term

wald.test.stat

#wald p-value
2 * pnorm(-abs(wald.test.stat))

```


